---
title: "NYPD - Shootings Data"
output: html_document
---

# Introduction

The NYPD Shooting Incidents data set includes shooting incident data from January 1st, 2006 through December 31st, 2024.
As such, it very interestingly covers the Covid19 Pandemic period.

The question I have about this data set is whether time of day impacts shooting incidents and whether the total number
of shootings are trending up or down over time.

```{r setup, include=FALSE}
library(rmarkdown)
library(knitr)
knitr::opts_chunk$set(echo = TRUE)

# tidy
library(tidyverse)
library(lubridate)
library(hms)
```

```{r covid19}
url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"

file_names <- c(
  "time_series_covid19_confirmed_global.csv",
  "time_series_covid19_deaths_global.csv",
  "time_series_covid19_confirmed_US.csv",
  "time_series_covid19_deaths_US.csv"
)

urls <- str_c(url_in, file_names)

global_cases <- read_csv(urls[1])
global_deaths <- read_csv(urls[2])
us_cases <- read_csv(urls[3])
us_deaths <- read_csv(urls[4])
```

```{r dataset import}
nypd_shootings <- read_csv("https://data.cityofnewyork.us/api/v3/views/833y-fsy8/export.csv")

spec(nypd_shootings)
```

```{r dataset tidy}
nypd_shootings <- nypd_shootings %>%
  select(c(INCIDENT_KEY, OCCUR_DATE, OCCUR_TIME)) %>%
  mutate(
    OCCUR_DATE = mdy(OCCUR_DATE),
    
    year = year(OCCUR_DATE),
    hour = hour(as_hms(OCCUR_TIME))
  )

nypd_shootings
tail(nypd_shootings)
```

```{r shootings by time of day}
nypd_shootings %>%
  distinct(INCIDENT_KEY, hour) %>%
  count(hour) %>%
  ggplot(aes(hour, n)) +
  theme_minimal() +
  geom_col(fill = "purple") +
  scale_x_continuous(breaks = 0:23) +
  labs(
    title = "NYPD Shootings by Time of Day",
    x = "Hour",
    y = "Number of incidents"
  )
```

```{r covid19 dates}
us_cases

covid_dates <- us_cases %>%
  pivot_longer(
    cols = matches("^\\d{1,2}/\\d{1,2}/\\d{2}$"),
    names_to = "date",
    values_to = "cases"
  ) %>%
  #filter(cases > 300) %>%
  mutate(date = mdy(date)) %>%
  distinct(date)

covid_dates

covid_start <- min(covid_dates$date)
covid_end <- max(covid_dates$date)

numeric_covid_start <- year(covid_start) + (month(covid_start) - 1) / 12
numeric_covid_end <- year(covid_end) + (month(covid_start) - 1) / 12

covid_start
covid_end
```

```{r shootings and covid19}
yearly_incidents <- nypd_shootings %>%
  distinct(INCIDENT_KEY, year) %>%
  count(year)

yearly_incidents %>%
  ggplot(aes(year, n)) +
  theme_minimal() +
  
  annotate(
    "rect",
    xmin = numeric_covid_start,
    xmax = numeric_covid_end,
    ymin = -Inf,
    ymax = Inf,
    fill = "lightblue",
    alpha = 0.3
  ) +
  
  geom_line() +
  geom_smooth(method = "lm", se = FALSE) +
  
  labs(
    title = "NYC Shootings by Year",
    subtitle = "Shaded area represents the COVID-19 pandemic period",
    x = "Year",
    y = "Number of incidents"
  )
```

```{r shootings by time of day during covid19}
nypd_shootings

nypd_shootings %>%
  filter(between(OCCUR_DATE, covid_start, covid_end)) %>%
  distinct(INCIDENT_KEY, hour) %>%
  count(hour) %>%
  ggplot(aes(hour, n)) +
  theme_minimal() +
  geom_col(fill = "purple") +
  scale_x_continuous(breaks = 0:23) +
  labs(
    title = "NYPD Shootings by Time of Day During Covid19",
    x = "Hour",
    y = "Number of incidents"
  )
```

## Conclusion
I was very interested in the timing of shootings from this data set, because I wanted to be able to draw
conclusions from the data. However, the more I think about this, the more I find that the data may be influenced by biases
that I have and that may be present within the data itself.

For example, I notice that time of day seems to play a major role in the number of shooting incidents. Incidents
late at night immediately draw my mind to thoughts of the bar scene when there is no data to back up any correlation
to population at bars and the subset of that population involved in shooting incidents.

It is also possible that the source of data is already biased. For example, if more police are in areas where
shooting incidents concentrate in the evenings and early morning, it could contaminate the data set.

What's also interesting to me about this data set is that it seems to show an overlapping increase with the Covid19 data set from the
course. During the Covid19 pandemic, it appears that recorded shooting incidents rose in 2020 and 2021.

Overall, it does appear that shooting incidents are trending down, and the time of day that shooting incidents reportedly
occurred roughly doubled in the evening and early morning hours.

```{r session}
sessionInfo()
```
